Figures

### 2016-08-16 ms_3_res_disc_concl.Rmd
cleaning out old plots; updating figures; editing manuscript

The IUCN publishes species range maps based on expert input of spatial boundaries of a given species' "limits of distribution" [@iucn_iucn_2015] - essentially a refined extent of occurrence, based on observation records and informed by expert understanding of species' range and habitat preferences.  In contrast, AquaMaps models species distribution based on environmental preferences (e.g., temperature, depth, salinity) deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a 0.5 degree grid, creating a global map of "probability of occurrence" for each species [@kaschner_mapping_2006; @ready_predicting_2010].  Studies using AquaMaps data frequently define "presence" by applying a probability threshold, e.g., the Ocean Health Index defines presence as 40% or greater probability of occurrence [@halpern_index_2012; @halpern_patterns_2015].

The methodologies behind the creation of these datasets imply differences in prediction of species distribution due to errors of commission (falsely indicating species presence) and omission (falsely indicating species absence).  Geographic range data such as IUCN range maps frequently introduce commission errors, while species distribution models such as AquaMaps will likely introduce fewer commission errors but more omission errors.  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range [@rondinini_tradeoffs_2006].  By comparing maps resulting from IUCN and AquaMaps methodologies, we can identify and possibly address mechanistic causes for each type of error.

The two datasets have notably different taxonomic (Fig. 1A) and regional (Figs. 1B, 1C) coverage.  AquaMaps encompasses a broader range of taxa than IUCN, as IUCN spatial data files are only available for select taxonomic groups that have been comprehensively assessed. While species numbers in both datasets peak in tropical latitudes near the equator, species counts for IUCN maps drop quickly beyond 30°N and 30°S, while species counts for AquaMaps remain robust well into temperate latitudes.  The longitude frequency plots show a slight shift in the IUCN dataset away from the Atlantic and eastern Pacific compared to AquaMaps.  To achieve more comprehensive global coverage of species ranges these two datasets can be used together, but the underlying methodological differences complicate such direct comparisons.

``` {r load_spp_list_by_quad, echo = FALSE}
spp_list_quads <- read.csv('data/spp_list_quads.csv', stringsAsFactors = FALSE)
n_q1 <- nrow(spp_list_quads %>% filter(quad == 'q1'))
n_q2 <- nrow(spp_list_quads %>% filter(quad == 'q2'))
# n_q2_am_sm  <- nrow(spp_list_quads %>% filter(sm_range == 'AM' & quad == 'q2'))
n_q2_corals <- nrow(spp_list_quads %>% filter(str_detect(tolower(spp_group), 'coral') & quad == 'q2'))
n_corals    <- nrow(spp_list_quads %>% filter(str_detect(tolower(spp_group), 'coral')))
n_q3 <- nrow(spp_list_quads %>% filter(quad == 'q3'))
n_q4 <- nrow(spp_list_quads %>% filter(quad == 'q4'))

@@

To explore differences in species distribution and range between the two datasets, we plotted the distribution alignment (how much of the smaller range falls within the larger range, i.e., where on the map) against the area alignment (ratio of smaller range area to larger range area, i.e., how much of the map) for each shared species (Fig. 2A). 
 This analysis revealed a general negative linear pattern, suggesting that increasing similarity in range area correlates with decreasing distribution alignment.
 AquaMaps tends to extrapolate species ranges into suitable areas beyond known occurrences, such that each additional unit of range predicted by AquaMaps will fall in different locations than an additional unit of range predicted using IUCN methodology. For species with dissimilar range areas, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar range areas, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

The mean distribution alignment for species included in both datasets was `r dist_align_mean`%; the mean area alignment was `r area_align_mean`%. By dividing the map-paired species into quadrants based on these means, we highlight categories of relationships that help further explain this general pattern. Representative maps from each category are provided in the supporting materials (Fig. S1).

The upper right quadrant includes species (n = `r n_q1`) whose described ranges are above average in alignment of both spatial distribution and area. These species tend to be well-studied and include wide-ranging pelagic organisms such as marine mammals, tunas, and billfishes (Fig. 2B). This result is not surprising, as species with very large ranges are likely to be more aligned regardless of methodology simply because their ranges span nearly the entire map.

The area-mismatched ranges contained in the upper left quadrant (n = `r n_q2`) include many species whose spatial distribution is similar, but where the IUCN range is notably larger, often extending into deeper water.  For example, corals dominate this quadrant (n = `r n_q2_corals`; `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in this quadrant), and IUCN range maps tend to extend corals into waters beyond their preferred depths, likely introducing errors of commission  (Fig. S3). 
 Ocean depth preference is explicitly included in AquaMaps models, while depth is frequently overlooked as a factor in IUCN range maps.  Simply clipping IUCN range maps to known depth preferences would resolve many of these mismatches.

Species found in the lower right quadrant (n = `r n_q3`) often represent cases of "two wrongs make a right."  For these species, IUCN ranges frequently overextend into unsuitable depths, as in the case of many upper left quadrant species, while at the same time AquaMaps ranges often aggressively extrapolate presence into locations where IUCN predicts absence, introducing additional commission errors.  Consequently, area ratios are close to 100%, though similar areas are unhelpful when the distributions are poorly aligned.

The most vexing cases are in the lower left quadrant (n = `r n_q4`), where neither distribution nor area match well. Data-poor species are more common in this quadrant; indeed, the median number of species occurrence records (averaging occurrences from the Ocean Biogeographic Information System (OBIS) [@obis_data_2016] and the Global Biodiversity Information Facility (GBIF) [@gbif_global_2010]) for this quadrant is `r med_occ_q4` records, compared to a median of `r med_occ_q123` records for species across the other three quadrants.  When extrapolating from limited observations, the AquaMaps model often predicts species presence well beyond known occurrences, introducing commission errors; at the same time, IUCN range maps generally target known occurrences, possibly introducing omission errors for data-limited species.

@@

Method-driven differences in commission and omission errors produce clear and significant disagreement in species range descriptions between AquaMaps and IUCN datasets.   To examine the implications of these differences, we repeated two recent studies, varying only the prioritization of one data set over the other.

@@

The global Ocean Health Index (OHI) [@halpern_index_2012; @halpern_patterns_2015], a composite index comprising ten sustainable benefits provided by a healthy ocean, uses species spatial distribution data and IUCN Red List conservation status to calculate biodiversity status (scored from zero to 100) for each of the world's 221 exclusive economic zones. To maximize the number of represented species, OHI gleans spatial distribution data from Red List-assessed species in both IUCN and AquaMaps datasets (n = `r format(n_spp_ohi, big.mark = ',')`), prioritizing IUCN data for the `r format(n_spp_both_ohi, big.mark = ',')` species included in both sources.  OHI currently uses a probability of occurrence threshold of 40% to determine species presence for AquaMaps data.

We calculated the OHI species status score under several scenarios to observe the impact of toggling the prioritized data set from IUCN to AquaMaps, and toggling the AquaMaps presence threshold from 40% to 0% (<a href = '#fig3'>Fig. 3</a>).  Reducing the threshold increases the apparent range of a species, so a decrease in a region’s score under scenario 1 would indicate increased spatial representation of threatened species for that region.  Shifting priority from IUCN to AquaMaps may increase or decrease a species’ apparent range, so a decrease in region score for scenario 2 may indicate a decrease in spatial representation of low-risk species, an increase in spatial representation of high-risk species, or a combination of both. Scenario three combines effects of scenarios 1 and 2.

Given that only `r p_spp_both_ohi`% of species are included in both datasets, it is surprising that changing the priority for overlapping species from IUCN maps to AquaMaps would result in substantial country-level score shifts as seen in scenarios 2 and 3.  While the mean global score did not vary significantly from scenario to scenario, select countries gained up to 7.5 points while others dropped as many as 5 points.  This result indicates that especially on a national or regional scale, an arbitrary change in how the two datasets are combined can result in a different assessment of species conservation status.

@@

AquaMaps and IUCN range maps show reasonable agreement for many well-studied species, but substantial differences illustrate uncertainty in our understanding of spatial distribution for many others.  Although many other approaches exist for species distribution modeling, these two are the only ones applied globally across different marine taxa.  Identifying and addressing differences in these datasets will increase their utility for research and conservation actions.  Several likely drivers of commission and omission errors between these datasets point to a few important ways to improve range alignment.

For IUCN range data, clipping ranges to known depth limits improves output for many species, most notably corals and reef-associated fishes.  If species' depth limits are not known, simple rules of thumb will likely reduce commission errors without introducing substantial omission errors.  For example, for most corals, researchers could clip range maps to the photosynthetic limit of 200 meters.
 For AquaMaps range data, dependent primarily on environmental and physical preferences and conditions, implementing area restrictions based on biogeographical criteria such as Marine Ecoregions of the World [@spalding_marine_2007] would likely decrease commission errors and improve predictive power, especially for data-poor species.
 For either data set, maps based on few occurrences are more likely to include errors of commission and omission.  Occurrence counts from external sources such as OBIS or GBIF can help identify relatively data-poor species; additionally, the "occurcells" attribute in the AquaMaps data set, which counts the number of half-degree cells used to generate the environmental envelope for each species, can be used in a similar manner.

Many studies using AquaMaps data test the sensitivity of results by varying the probability of occurrence threshold to determine presence.  This decision ultimately represents a tradeoff between errors of commission (low threshold) and omission (high threshold).  Using AquaMaps in conjunction with IUCN can mitigate potential errors, while also increasing the taxonomic and spatial breadth of coverage, as long as the differences between the datasets can be reasonably minimized. In this case, we recommend a presence threshold of 0% as it most closely approximates the "limits of distribution" criterion defined by IUCN data providers.

>>>>

(R) The two datasets have notably different taxonomic (Fig. 1A) and regional (Figs. 1B, 1C) coverage.  AquaMaps encompasses a broader range of taxa than IUCN, as IUCN spatial data files are only available for select taxonomic groups that have been comprehensively assessed. While species numbers in both datasets peak in tropical latitudes near the equator, species counts for IUCN maps drop quickly beyond 30°N and 30°S, while species counts for AquaMaps remain robust well into temperate latitudes.  The longitude frequency plots show a slight shift in the IUCN dataset away from the Atlantic and eastern Pacific compared to AquaMaps.

(D) _To achieve more comprehensive global coverage of species ranges these two datasets can be used together, understanding that the underlying differences complicate such direct comparisons.  Major considerations include aligning the criteria used to determine "presence" (e.g. selecting an appropriate presence threshold for AquaMaps) and reducing mechanisms that introduce errors of commission and omission._

``` {r load_spp_list_by_quad, echo = FALSE}
spp_list_quads <- read.csv('data/spp_list_quads.csv', stringsAsFactors = FALSE)
n_q1 <- nrow(spp_list_quads %>% filter(quad == 'q1'))
n_q2 <- nrow(spp_list_quads %>% filter(quad == 'q2'))
n_q2_iucn_larger <- nrow(spp_list_quads %>% filter(sm_range == 'AM' & quad == 'q2'))
n_q2_corals <- nrow(spp_list_quads %>% filter(str_detect(tolower(spp_group), 'coral') & quad == 'q2'))
n_corals    <- nrow(spp_list_quads %>% filter(str_detect(tolower(spp_group), 'coral')))
n_q3 <- nrow(spp_list_quads %>% filter(quad == 'q3'))
n_q4 <- nrow(spp_list_quads %>% filter(quad == 'q4'))

@@

___from Methods___
(D) _The methodologies behind the creation of these datasets imply differences in prediction of species distribution due to errors of commission (falsely indicating species presence) and omission (falsely indicating species absence).  Geographic range data such as IUCN range maps frequently introduce commission errors, while species distribution models such as AquaMaps will likely introduce fewer commission errors but more omission errors.  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range [@rondinini_tradeoffs_2006].  By comparing maps resulting from IUCN and AquaMaps methodologies, we can identify and possibly address mechanistic causes for each type of error._

(D/M) ___is this sufficiently described in methods? or revisit here?___ ___Both datasets strive for accuracy in different ways; the truth likely lies somewhere in between.___ To explore differences in species distribution and range between the two datasets, we plotted the distribution alignment (how much of the smaller range falls within the larger range, i.e., where on the map) against the area alignment (ratio of smaller range area to larger range area, i.e., how much of the map) for each shared species (Fig. 2A).  Where two independent methods predict presence (or absence) of a species in the same location, we can assume a higher confidence in the accuracy of that prediction.  Where one method predicts presence while the other predicts absence, we can investigate potential sources of commission and omission errors to refine our understanding of the species' range.  Understanding these sources of error can provide guidance to improve predictions of species range even when only one source of spatial data is available.

(R) This analysis revealed a general negative linear pattern, suggesting that increasing similarity in range area correlates with decreasing distribution alignment.
 AquaMaps tends to extrapolate species ranges into suitable areas beyond known occurrences, such that each additional unit of range predicted by AquaMaps will fall in different locations than an additional unit of range predicted using IUCN methodology. For species with dissimilar range areas, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar range areas, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

(R) The mean distribution alignment for species included in both datasets was `r dist_align_mean`%; the mean area alignment was `r area_align_mean`%. By dividing the paired map species into quadrants based on these means, we highlight categories of relationships that help further explain this general pattern. Representative maps from each category are provided in the supporting materials (Fig. S1).

(R) The upper right quadrant includes the species (n = `r n_q1`) whose described ranges are above average in alignment of both spatial distribution and area. These species tend to be well-studied and include wide-ranging pelagic organisms such as marine mammals, tunas, and billfishes (Fig. 2B). This result is not surprising, as species with very large ranges are likely to be more aligned regardless of methodology simply because their ranges span nearly the entire map.

(R) The area-mismatched ranges contained in the upper left quadrant (n = `r n_q2`) include many species whose spatial distribution is similar, but where one range is notably larger than the other (For `r round(n_q2_iucn_larger/n_q2 * 100, 1)`% of the species in this quadrant, the IUCN range is an average of 2.57 times larger than the AquaMaps range).  This suggests a high rate of commission and/or omission errors by one or both datasets; further analysis is required to disentangle the source or type of error contributing to poor area alignment.  The results of the coral analysis (described below) provide some insight.

(R) ___area ratio suggests similar commission/omission errors, but dist suggests otherwise___ Species found in the lower right quadrant (n = `r n_q3`) seem to represent cases of "two wrongs make a right."  For these species, IUCN and AquaMaps both predict ranges extending far beyond the overlapping region, but the methodological differences result in very different extrapolations.   ___OMIT:___ _For these species, IUCN ranges frequently overextend into unsuitable depths, as in the case of many upper left quadrant species, while at the same time AquaMaps ranges often aggressively extrapolate presence into locations where IUCN predicts absence, likely introducing additional commission errors._  Consequently, area ratios are close to 100%, though the poor distribution alignment indicates that one or both datasets are introducing errors. As above, further analysis is required to disentangle the causes of error.

(R) The lower left quadrant includes species (n = `r n_q4`) where alignment is poor in both dimensions.  ___present a solution here?___ Data-poor species are more common in this quadrant; indeed, the median number of species occurrence records (averaging occurrences from the Ocean Biogeographic Information System (OBIS) [@obis_data_2016] and the Global Biodiversity Information Facility (GBIF) [@gbif_global_2010]) for this quadrant is `r med_occ_q4` records, compared to a median of `r med_occ_q123` records for species across the other three quadrants.  It is not surprising that range maps based upon fewer observations bear greater uncertainty.  ___OMIT:___ _When extrapolating from limited observations, the AquaMaps model often predicts species presence well beyond known occurrences, introducing commission errors; at the same time, IUCN range maps generally target known occurrences, possibly introducing omission errors for data-limited species._

## New analysis: Coral maps (Distribution-aligned quads)

(D/M) Noting from Fig. 2B that corals dominate the upper-left "distribution-aligned" quadrant of Fig. 2A (n = `r n_q2_corals`; `r round(n_q2_corals/n_q2 * 100, 1)`% of all species in this quadrant), we clipped the IUCN coral range maps to a 200 m bathymetry raster to examine the impact on distribution and area alignment.  Ocean depth preference is explicitly included in AquaMaps models.  While depth is recommended by the IUCN as a criterion for providers of range maps ("The limits of distribution can be determined by using known occurrences of the species, along with the knowledge of habitat preferences, remaining suitable habitat, elevation limited, and other expert knowledge of the species and its range." [@REF http://www.iucnredlist.org/technical-documents/red-list-training/iucnspatialresources]), it is clear that this cannot be taken for granted.

- Coral area plot

(R) Fig. 3A shows the 463 coral species mapped in the IUCN dataset, with their ranges broken into proportional area deeper and shallower than 200 m.  None of these species is indicated to occur deeper than 200 m, and 94% are confined to waters shallower than 50 m; seven of the mapped species had no reported depth information.  Clipping coral ranges to shallower than 200 m eliminated an average of 47.6% of the total predicted area while still allowing for a conservative estimate of suitable habitat.

- Coral new quads

(R) Masking coral ranges to shallow waters reduces a clear source of commission errors.  In so doing, we see a strong increase in the apparent alignment of species maps between IUCN and AquaMaps.  Using the original quadrant definitions from Fig. 2B, we see in Fig. 3B and 3C a massive shift in 354 paired map species due to constraining coral depths to 200 m.  Membership in the "well-aligned" quadrant jumped from 22.4% to 76.2%, with a corresponding decrease in all other quadrants.

(D) We cannot know for certain the true distribution of each of these corals; much of the apparent over-prediction of coral ranges is likely due to very conservative and precautionary approach.  But a small and sensible shift in method drastically decreases the introduction of likely commission errors, greatly improving our confidence in the remaining reported ranges.  Note that this change applies just as well to the IUCN coral maps that are not included in the paired map analysis, and likely to other reef-associated flora and fauna.  While species depth preferences are an easy and consistent means of constraining range predictions, other conditions such as salinity and temperature could be cautiously used to refine the results of expert opinions.

***What is missing here is something to show that AquaMaps ranges, clipped to FAO Major Fishing Areas, are probably introducing errors of commission as well*** - _the fish don't see that FAO boundary and suddenly decide not to cross; most obvious in the boundaries defined by longitudes.  Can we show this with a couple of example maps?  That wasn't received well for the corals, but I can't think of a clear and ambiguous way to analyze this problem._

@@

Neither IUCN nor AquaMaps claims a lock on the "truth", and the small overlap in mapped species means each provides value in taxonomic and geographic coverage.  Method-driven differences in commission and omission errors produce clear and significant disagreement in species range descriptions between AquaMaps and IUCN datasets.  To examine the implications of these differences, we repeated two recent studies, varying only the selection of one data set over the other.

@@

The global Ocean Health Index (OHI) [@halpern_index_2012; @halpern_patterns_2015], a composite index comprising ten sustainable benefits provided by a healthy ocean, uses species spatial distribution data and IUCN Red List conservation status to calculate biodiversity status (scored from zero to 100) for each of the world's 221 exclusive economic zones. To maximize the number of represented species, OHI gleans spatial distribution data from Red List-assessed species in both IUCN and AquaMaps datasets (n = `r format(n_spp_ohi, big.mark = ',')`), prioritizing IUCN data for the `r format(n_spp_both_ohi, big.mark = ',')` species included in both sources.

___OMIT:___ _We calculated the OHI species status score under several scenarios to observe the impact of toggling the prioritized data set from IUCN to AquaMaps, and toggling the AquaMaps presence threshold from 40% to 0% (<a href = '#fig3'>Fig. 3</a>).  Reducing the threshold increases the apparent range of a species, so a decrease in a region’s score under scenario 1 would indicate increased spatial representation of threatened species for that region.  Shifting priority from IUCN to AquaMaps may increase or decrease a species’ apparent range, so a decrease in region score for scenario 2 may indicate a decrease in spatial representation of low-risk species, an increase in spatial representation of high-risk species, or a combination of both. Scenario three combines effects of scenarios 1 and 2._

___OMIT:___ _Given that only `r p_spp_both_ohi`% of Red List-assessed species are included in both datasets, it is surprising that changing the priority for overlapping species from IUCN maps to AquaMaps would result in substantial country-level score shifts as seen in scenarios 2 and 3.  While the mean global score did not vary significantly from scenario to scenario, select countries gained up to 7.5 points while others dropped as many as 5 points.  This result indicates that especially on a national or regional scale, an arbitrary change in how the two datasets are combined can result in a different assessment of species conservation status._

@@

AquaMaps and IUCN range maps show reasonable agreement for many well-studied species, but substantial differences illustrate uncertainty in our understanding of spatial distribution for many others.  Although many other approaches exist for species distribution modeling, these two are the largest approaches applied globally across a broad range of marine taxa.  Identifying and addressing differences in these datasets will increase their utility for research and conservation actions.

For IUCN range data, clipping ranges to known depth limits improves confidence in predicted ranges for coral species; similarly, depth may be a useful characteristic to consider for other reef-associated organisms.  Other parameters such as salinity and temperature may provide useful constraints on other taxa.  Simple but conservative rules of thumb will likely reduce commission errors without introducing substantial omission errors.
 For AquaMaps range data, dependent primarily on environmental and physical preferences and conditions, implementing area restrictions based on biogeographical criteria such as Marine Ecoregions of the World [@spalding_marine_2007], rather than political and economic criteria such as FAO Major Fishing Areas, would likely decrease commission errors and improve predictive power, especially for data-poor species.
 For either data set, maps based on few occurrences are more likely to bear high uncertainty in range predictions.  Occurrence counts from external sources such as OBIS or GBIF can help identify relatively data-poor species; additionally, the "occurcells" attribute in the AquaMaps data set, which counts the number of half-degree cells used to generate the environmental envelope for each species, can be used in a similar manner.

___OMIT:___ _Many studies using AquaMaps data test the sensitivity of results by varying the probability of occurrence threshold to determine presence.  This decision ultimately represents a tradeoff between errors of commission (low threshold) and omission (high threshold).  Using AquaMaps in conjunction with IUCN can mitigate potential errors, while also increasing the taxonomic and spatial breadth of coverage, as long as the differences between the datasets can be reasonably minimized.  In this case, a presence threshold of 0% most closely approximates the "limits of distribution" criterion defined by IUCN data providers._

### 2016-08-16 ms_5_figures.Rmd
cleaning out old plots; updating figures; editing manuscript

### Figure 3

<a id = 'fig3'></a>

![paired map comparison](figures/fig3a_violinplot_paired_maps.png)

why is the global one not showing up?:
![global full species list comparison](figures/fig3a_violinplot_ohi_global.png)

@@

![Change in MPA coverage](figures/fig4_barchart_mpa_gap.png)

### New figures - corals

![coral area change due to depth clipping](figures/fig5a_coral_area_plot.png)

![coral quadmap change due to depth clipping](figures/fig5b_scatter_quadmap_corals.png)

>>>>

### Fig. 3
<a id = 'fig3'></a>

![coral area change due to depth clipping](figures/fig3a_coral_area_plot.png)

![coral quadmap change due to depth clipping](figures/fig3b_scatter_quadmap_corals.png)

![coral quadmap breakdown](figures/fig3c_barchart_coral_quads.png)

@@

![paired map comparison](figures/fig4a_violinplot_paired_maps.png)

why is the global one not showing up?:
![global full species list comparison](figures/fig4a_violinplot_ohi_global.png)

### Figure 5

<a id = 'fig5'></a>

![Change in MPA coverage](figures/fig5_barchart_mpa_gap.png)
